---
title: "Moneyball Multiple Regression Model"
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    keep_md: yes
    theme: cerulean
    toc: yes
  pdf_document: default
---

```{r load-libaries, eval=TRUE, include=FALSE}
library(fBasics)
library(knitr)
```

```{r load-data, eval=TRUE,include=FALSE}
setwd("~/IS621Summer2016/HW1")
BB.list <- read.csv("more/moneyball-training-data.csv",
    stringsAsFactors=FALSE, sep=",")
BB.df <- data.frame(BB.list)
DescStats.df <- basicStats(BB.df)
```

# Data Dictionary

variable            | definition                              | effect
----------------    | -----------                             |------------
`INDEX`             | Identification Variable (do not use)    | None
`TARGET_WINS`       | Number of wins                          | Positive
`TEAM_BATTING_H`    | Base Hits by batters (1B,2B,3B,HR)      | Positive
`TEAM_BATTING_2B`   | Doubles by batters (2B)                 | Positive
`TEAM_BATTING_3B`   | Triples by batters (3B)                 | Positive
`TEAM_BATTING_HR`   | Homeruns by batters (4B)                | Positive
`TEAM_BATTING_BB`   | Walks by batters                        | Positive
`TEAM_BATTING_SO`   | Strikeouts by batters                   | Negative
`TEAM_BATTING_HBP`  | Batters hit by pitch (get a free base)  | Positive
`TEAM_BASERUN_SB`   | Stolen bases                            | Positive
`TEAM_BASERUN_CS`   | Caught stealing                         | Negative
`TEAM_PITCHING_H`   | Hits allowed                            | Negative
`TEAM_PITCHING_HR`  | Homeruns allowed                        | Negative
`TEAM_PITCHING_BB`  | Walks allowed                           | Negative
`TEAM_PITCHING_SO`  | Strikeouts by pitchers                  | Positive
`TEAM_FIELDING_E`   | Errors                                  | Negative
`TEAM_FIELDING_DP`  | Double Plays                            | Positive
----------------    | -----------                             |------------

# Exploring the data

Many of the statistics in the data sets provided have been extrapolated using base statistics from the deadball 
era circa ~1900-1920 and prior.  Outliers that need to be adjusted can be found using the reference link [!baseball-almanac](http://www.baseball-almanac.com/recbooks/rb_stba2.shtml).  Note that during the deadball
era, a nearly soft ball was used which had dramatic effect on power hitting and pitching statistics.
Any adjustment that rationalize the data from this period into observations that include post WWII statistics
should be bound by the later era's limits so distributions aren't skewed.

```{r hist-wins, eval = TRUE,include=FALSE}
hist(BB.df$TARGET_WINS)
```

Generalized Equation for Multiple Regression
\[
\begin{aligned}
\widehat{wins} &= \hat{\beta}_0 + 
                    \hat{\beta}_1 \times hits  + 
                    \hat{\beta}_2 \times doubles + 
                    \hat{\beta}_3 \times triples +
                    \hat{\beta}_4 \times homeruns + 
                    \hat{\beta}_5 \times walks + 
                    \hat{\beta}_6 \times strikeouts . . . +
                    \end{aligned}
\]

## Full Panel
```{r full, eval = TRUE}
#wins x full panel
pairs(TARGET_WINS~.,
        data=BB.df,pch=".",gap=.5,upper.panel=panel.smooth)
ml.full <- lm(TARGET_WINS~.
                  , data = BB.df)
summary(ml.full)
```

## Slugging
The most hits in a single season (unadjusted) is 1783--NL Philadelphia Phillies 1930.
The most doubles in a single season (unadjusted) is 376--AL Texas in 2008.
The most triples in a single season (unadjusted) is 153--NL Baltimore in 1894.
Records that exceed these amounts should be adjusted either to NA or the median.

```{r summary-slugging, eval=TRUE}
kable(DescStats.df[,2:6],align='l',digits = c(0,2,2))
```

```{r boxplot-slugging, eval = TRUE,include=FALSE}
boxplot(BB.df[,3:6],main="Baseball Team Stats", 
        	xlab="Slugging") 
```

```{r slugging, eval = TRUE, tidy = TRUE}
#wins x batting - slugging
pairs(TARGET_WINS~TEAM_BATTING_H+TEAM_BATTING_2B+TEAM_BATTING_3B+TEAM_BATTING_HR,
        data=BB.df,pch=".",gap=.5,upper.panel=panel.smooth)
ml.slugging <- lm(TARGET_WINS~TEAM_BATTING_H+TEAM_BATTING_2B+TEAM_BATTING_3B+TEAM_BATTING_HR
                  , data = BB.df)
summary(ml.slugging)
```

## On-Base
102 strikeout NAs can remain as long as they're not counted in descriptive statistics 
as observations.  2085 hit by pitch NAs disqualifies this field from use in the model.

The most walks in a single season (unadjusted) is 835--AL Boston Red Sox in 1949.
The fewest walks in a single season (unadjusted) is 282--NL St. Louis Cardinals 1908.
Records that exceed these amounts should be adjusted either to NA or the median.

```{r summary-onbase, eval=TRUE}
kable(DescStats.df[,c(7,8,11)],align='l',digits = c(0,2,2))
```

```{r boxplot-onbase, eval = TRUE}
boxplot(BB.df[,c(7,8,11)],main="Baseball Team Stats", 
        	xlab="On Base") 
```

```{r onbase, eval = TRUE, tidy = TRUE}
#wins x batting - on-base
pairs(TARGET_WINS~TEAM_BATTING_BB+TEAM_BATTING_SO+TEAM_BATTING_HBP,
        data=BB.df,pch=".",gap=.5,upper.panel=panel.smooth)
ml.onbase <- lm(TARGET_WINS~TEAM_BATTING_BB+TEAM_BATTING_SO+TEAM_BATTING_HBP
                  , data = BB.df)
summary(ml.onbase)
```

## Base Running
131 stolen base NAs can & 772 caught stealing NAs can remain as long as they're not counted
in descriptive statistics.

The most stolen bases in a single season (unadjusted) is 426--NL New York in 1893.
The most caught stealing bases count in a single season (unadjusted) is 191--AL NY in 1914.
Records that exceed these amounts should be adjusted either to NA or the median.

These statistics are co-linear and may be better used as a derived statistic for
Expected value of team stolen bases E(SB) = SB * likelihood of success (SB/SB attempts)

```{r summary-baserunning, eval=TRUE}
kable(DescStats.df[,9:10],align='l',digits = c(0,2,2))
```

```{r boxplot-baserunning, eval = TRUE,include=FALSE}
boxplot(BB.df[,9:10],main="Baseball Team Stats", 
        	xlab="Base Running") 
```

```{r baserunning, eval = TRUE, tidy = TRUE}
#wins x batting - base running
pairs(TARGET_WINS~TEAM_BASERUN_SB+TEAM_BASERUN_CS,
        data=BB.df,pch=".",gap=.5,upper.panel=panel.smooth)
ml.baserunning <- lm(TARGET_WINS~TEAM_BASERUN_SB+TEAM_BASERUN_CS
                  , data = BB.df)
summary(ml.baserunning)
```

## Fielding
286 Fielding error NAs can remain as long as they're not counted in descriptive statistics 
as observations.  

The most fielding errors in a single season (unadjusted) is 639--NL Philadelphia in 1883.
The most fielding errors in a single season (unadjusted) post WWII is 234--NL Philadelphia in 1945.
Records that exceed the post WWII amount should be adjusted either to NA or the median.

```{r summary-fielding, eval=TRUE}
kable(DescStats.df[,16:17],align='l',digits = c(0,2,2))
```

```{r boxplot-fielding, eval = TRUE}
boxplot(BB.df[,16:17],main="Baseball Team Stats", 
        	xlab="Fielding") 
```

```{r fielding, eval = TRUE, tidy = TRUE}
#wins x batting - base running
pairs(TARGET_WINS~TEAM_FIELDING_E+TEAM_FIELDING_DP,
        data=BB.df,pch=".",gap=.5,upper.panel=panel.smooth)
ml.fielding <- lm(TARGET_WINS~TEAM_FIELDING_E+TEAM_FIELDING_DP
                  , data = BB.df)
summary(ml.fielding)
```

## Pitching
286 Strikeout error NAs can remain as long as they're not counted in descriptive statistics 
as observations.  

The most hits given up in a single season (unadjusted) that I could find was fewer that 2000.
The most homeruns given up in a single season (unadjusted) that I could find was fewer that 250.
The most walks given up in a single season (unadjusted) that I could find was fewer that 800.
The most strikeouts thrown in a single season (unadjusted) that I could find was fewer that 1400.
Records that exceed these amount should be adjusted either to NA or the median.

```{r summary-pitching, eval=TRUE}
kable(DescStats.df[,12:15],align='l',digits = c(0,2,2))
```

```{r boxplot-pitching, eval = TRUE}
boxplot(BB.df[,12:15],main="Baseball Team Stats", 
        	xlab="Pitching") 
```

```{r pitching, eval = TRUE, tidy = TRUE}
#wins x batting - pitching
pairs(TARGET_WINS~TEAM_PITCHING_H+TEAM_PITCHING_HR+TEAM_PITCHING_BB+TEAM_PITCHING_SO,
        data=BB.df,pch=".",gap=.5,upper.panel=panel.smooth)
ml.pitching <- lm(TARGET_WINS~TEAM_PITCHING_H+TEAM_PITCHING_HR+TEAM_PITCHING_BB+TEAM_PITCHING_SO
                  , data = BB.df)
summary(ml.pitching)
```


